---
title: "Graphs"
output: html_document
---

```{r setup, include=FALSE, message=FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE)  
# Global chunk option:
# settings defined here apply to all subsequent chunks unless overridden.
# e.g echo = FALSE hides the R code and shows only the output/results. for all chunks
# message = FALSE , to avoid package startup messages

if (!require(pacman))  # checks if pacman is installed
  install.packages("pacman") # if pacman doesnt exist on this machine install it
suppressPackageStartupMessages({pacman::p_load( # installs and loads packages
  tidyverse, 
  inspectdf,
  plotly,
  janitor,
  visdat,
  esquisse,
  here,
  readxl,
  flextable,
  gt,
  reactable,
  lubridate,
  Sleuth3,
  Sleuth2,
  reshape,
  glue,
  MASS
   )})
```

# To make graphs with ggplot2, the data must be in a data frame

```{r echo = FALSE, warning=FALSE}
        # use FACTOR so as to define the order of the categories
        # Without factor ordering, categories may appear alphabetically.
dat <- data.frame(
    time = factor(c("Lunch","Dinner"), levels=c("Lunch","Dinner")), 
    total_bill = c(14.89, 17.23)
)

# BAR GRAPHS OF VALUES 

# Very basic bar graph
# Assign different fill colors based on the categories in time.
# Add a black outline on the bars
# the height of the bar will represent the value in a column of the data frame , by using stat="identity"

ggplot(data=dat, aes(x=time, y=total_bill, fill=time)) +
    geom_bar(colour="black", width=.3,stat="identity", show.legend = FALSE) +  # No legend 
    #guides(fill="none") + # No legend 
    ggtitle("Average bill for 2 people")


# BAR GRAPHS OF COUNTS

# the height of the bar will represent the count of cases. This is done by using stat="bin" (which is the default).
# Will use the TIPS data from RESHAPE Package


ggplot(data=tips, aes(x=day, fill = day)) +
    geom_bar(colour="black", width=.6, stat="count") + # same as geom_bar(colour="black", width=.8) since stat="bin" is the default
    guides(fill="none") + # No legend
    xlab("Day of the Week") + ylab("Total Count") 


# Basic line graph

# Change color of both line and points
# Change line type and point type, and use thicker line and larger points
# Change points to circles with black fill
# the data points must be grouped so that it knows which points to connect.All points should be connected, so group=1
ggplot(data=dat, aes(x=time, y=total_bill, group=1)) + 
    geom_line(colour="red", linetype="dashed", size=1.3) + 
    geom_point(colour="red", size=4, shape=21, fill="black")+ 
    expand_limits(y=0) +
    xlab("Time of day") + ylab("Total bill") +
    ggtitle("Average bill for 2 people")


```

# GRAPHS WITH MORE VARIABLES

```{r echo = FALSE, warning=FALSE}

dat1 <- data.frame(
    sex = factor(c("Female","Female","Male","Male"), levels = c("Female","Male")),
    time = factor(c("Lunch","Dinner","Lunch","Dinner"), levels=c("Lunch","Dinner")),
    total_bill = c(13.53, 16.81, 16.24, 17.42)
)



# BAR GRAPHS

# Bar graph, time on x-axis, color fill grouped by sex 
# use position_dodge() to tell ggplot “Place bars for different groups side-by-side instead of stacking them"
# color fill  GROUPED BY TIME

ggplot(data=dat1, aes(x=time, y=total_bill, fill=sex)) +
    geom_bar(color = "black",width =.6,stat="identity", position=position_dodge()) + 
    scale_fill_hue(name="Sex of payer") +      # Set legend title 
    xlab("Time of day") + ylab("Total bill") +
    ggtitle("Average bill for 2 people") 

# Color fill GROUPED BY SEX

ggplot(data=dat1, aes(x=sex, y=total_bill, fill=time)) +
    geom_bar(color = "black",width =.6,stat="identity", position=position_dodge()) + 
    scale_fill_hue(name="Time of Day") +      # Set legend title
    xlab("Time of day") + ylab("Total bill") +
    ggtitle("Average bill for 2 people")


# LINE GRAPHS

ggplot(data=dat1, aes(x=time, y=total_bill, group=sex, shape=sex, colour=sex)) + 
    geom_line(aes(linetype=sex), size=1) +     # Set linetype by sex
    geom_point(size=3, fill="white") +         # Use larger points, fill with white
    expand_limits(y=0) +                       # Set y range to include 0
    scale_colour_hue(name="Sex of payer",      # Set legend title
                     l=30)  +                  # Use darker colors (lightness=30)
    scale_shape_manual(name="Sex of payer",
                       values=c(22,21)) +      # Use points with a fill color
    scale_linetype_discrete(name="Sex of payer") +
    xlab("Time of day") + ylab("Total bill") + # Set axis labels
    ggtitle("Average bill for 2 people") +     # Set title
    theme_bw() +
    theme(legend.position=c(.7, .4))           # Position legend inside
                                               # This must go after theme_bw

```

# Plotting MEANS and ERROR BARS 

```{r echo = FALSE, warning=FALSE}
#load dataset
data(ToothGrowth)

# set primary colors to use

OJ_Color <- "#BEBEBE"
VC_Color <- "#0000FF"

# Summarize the data
tgc <- ToothGrowth %>%
  group_by(supp, dose) %>%
  summarise(
    N = n(),
    mean_len = mean(len, na.rm = TRUE),
    sd_len = sd(len, na.rm = TRUE),
    se = sd_len / sqrt(N),
    .groups = "drop"
  )

# LINE graphs

# Standard error of the mean

# to prevent the errorbars that share the same point from overlapping, use POSITION_DODGE to move them horizontally apart
pd <- position_dodge(0.1) # move grouped elements slightly side-by-side

# Blue error bars - notice the mapping of 'group=supp' -- without it, the error bars won't be dodged! (moved apart)

ggplot(tgc, aes(x=dose, y=mean_len, colour=supp, group = supp)) + 
    theme_classic() +
    geom_errorbar(aes(ymin=mean_len-se, ymax=mean_len+se),colour=VC_Color, width=.1 , position = pd) +
    geom_line(position=pd) +
    geom_point(position=pd, size = 2) +
    labs(x=NULL, y="Tooth Length") + # remove the X-axis LABEL
    scale_colour_hue(name="Supplement type",    # Legend label, use darker colors
                     breaks=c("OJ", "VC"),
                     labels=c("Orange juice", "Ascorbic acid"),
                     l=40) +                    # Use darker colors, lightness=40
    ggtitle("The Effect of Vitamin C on\nTooth Growth in Guinea Pigs") +
    expand_limits(y=0) +                        # Expand y range
    scale_y_continuous(breaks=0:20*4) +         # Set tick scale every 4
    theme(legend.justification=c(1,0),
          legend.position=c(1,0))               # Position legend in bottom right


# BAR GRAPHS

# Use dose as a factor rather than numeric
tgc2 <- tgc
tgc2$dose <- factor(tgc2$dose)



ggplot(tgc2, aes(x=dose, y=mean_len, fill=supp)) + 
    theme_classic() +
    geom_bar(position=position_dodge(), stat="identity") +
    geom_errorbar(aes(ymin=mean_len-se, ymax=mean_len+se),colour="black", size=.3, width=.1, position=position_dodge(.9)) +
    labs(x=NULL, y="Tooth Length") + # remove the X-axis LABEL
    scale_fill_hue(name="Supplement type",    # Legend label, use darker colors
                     breaks=c("OJ", "VC"),
                     labels=c("Orange juice", "Ascorbic acid"),
                     l=40) +                    # Use darker colors, lightness=40
    ggtitle("The Effect of Vitamin C on\nTooth Growth in Guinea Pigs") +
    expand_limits(y=0) +                        # Expand y range
    scale_y_continuous(breaks=0:20*4)         # Set tick scale every 4
  

```

```{r echo = FALSE, warning=FALSE}
set.seed(1234)
dat <- data.frame(cond = factor(rep(c("A","B"), each=200)), 
                   rating = c(rnorm(200),rnorm(200, mean=.8)))

## Basic histogram from the vector "rating". Each bin is .5 wide.
# Draw with black outline, white fill
ggplot(dat, aes(x=rating,fill=cond)) +
    geom_histogram(binwidth=.5, colour="black",position="dodge") + # place grouped objects side-by-side instead of overlapping.
    geom_vline(aes(xintercept=mean(rating, na.rm=T)),   # Add a line for mean and Ignore NA values for mean
               color="brown", linetype="dashed", size=1)

# Density curve
ggplot(dat, aes(x=rating,fill=cond)) + 
  geom_density(alpha=.3) + # alpha determines how opaque or transparent plot elements are.
  ggtitle("Distribution per Team") +
      theme(plot.title = element_text(lineheight=.8, face="bold")) + # Reduce line spacing and use bold text
  scale_fill_hue(name="Team Name",    # Legend label, use darker colors
                     breaks=c("A", "B"),
                     labels=c("LIVERPOOL", "MADRID"),
                     l=40)                    # Use darker colors, lightness=40

# FACETS , split the graphs based on a variable

ggplot(dat, aes(x=rating,fill=cond)) + 
  geom_density(alpha=.3) + # alpha determines how opaque or transparent plot elements are.
  ggtitle("Distribution per Team") +
  facet_grid(. ~ cond) +
      theme(plot.title = element_text(lineheight=.8, face="bold")) + # Reduce line spacing and use bold text
  scale_fill_hue(name="Team Name",    # Legend label, use darker colors
                     breaks=c("A", "B"),
                     labels=c("LIVERPOOL", "MADRID"),
                     l=40)   
```

# Exercises

```{r echo = FALSE, warning=FALSE}
# Examine the dataset trees, which should already be pre-loaded. Look at the help file using ?trees for more information about this data set. We wish to build a scatter plot that compares the height and girth of these cherry trees to the volume of lumber that was produced.
# 
# Create a graph using ggplot2 with Height on the x-axis, Volume on the y-axis, and Girth as the either the size of the data point or the color of the data point. Which do you think is a more intuitive representation?
# Add appropriate labels for the main title and the x and y axes.
# The R-squared value for a regression through these points is 0.36 and the p-value for the statistical significance of height is 0.00038. Add text labels “R-squared = 0.36” and “p-value = 0.0004” somewhere on the graph.

#load dataset
data(trees)
ggplot(trees,aes(x=Height,y=Volume, color = Girth)) +
  geom_point(color = "brown")+
  expand_limits(y=0) +
  ggtitle("Scatter plot for Cherry trees heights vs tree volume") +
  labs(x= "Tree Height", y= "Tree Volume") +
  annotate("label", x = min(trees$Height), y = max(trees$Volume) + 5
                           ,label = "R-squared = 0.36",color = "black", hjust = 0,fontface = "bold")  +
  annotate("label",x = min(trees$Height), y = max(trees$Volume) - 1
                           ,label = "P-value = 0.0004",color = "black",  hjust = 0,fontface = "bold")  +
  theme_classic() +
  theme(legend.position = 'none') # remove legend
  

```


```{r echo = FALSE, warning=FALSE}

Rosy <- data.frame(
  times = c(15, 11, 9, 12, 5, 2, 3),
  day   = 1:7,
  yhat  = c(14.36, 12.29, 10.21, 8.14, 6.07, 4.00,  1.93),
  lwr   = c( 9.54,  8.5,   7.22, 5.47, 3.08, 0.22, -2.89),
  upr   = c(19.18, 16.07, 13.2, 10.82, 9.06, 7.78,  6.75))

# Using ggplot() and geom_point(), create a scatterplot with day along the x-axis and times along the y-axis.

# Add a line to the graph where the x-values are the day values but now the y-values are the predicted values which we’ve called yhat. Notice that you have to set the aesthetic y=times for the points and y=yhat for the line. Because each geom_ will accept an aes() command, you can specify the y attribute to be different for different layers of the 

ggplot(Rosy,aes(x=day)) +
  geom_point(aes(y=times))+
  geom_line(aes(y=yhat), color = "brown") +
  expand_limits(y=0) +
  theme_classic() +
  theme(legend.position = 'none') # remove legend

# Add a ribbon that represents the confidence region of the regression line. The geom_ribbon() function requires an x, ymin, and ymax columns to be defined

ggplot(Rosy,aes(x=day)) +
geom_point(aes(y=times)) +
geom_line(aes(y=yhat), color = "blue") +
geom_ribbon( aes(ymin=lwr, ymax=upr), fill='forestgreen',alpha = 0.4) + # put some transparency of 0.4 to the band OR re-order and start with geom_ribbon first
expand_limits(y=0) +
ggtitle("Confidence Region Around the Regression Line") +
  labs(x= "Day of the week", y = "Time Played" ) +
theme_classic() +
theme(legend.position = 'none') # remove legend

```

```{r echo = FALSE, warning=FALSE}
data(birthwt, package='MASS')

# add better labels to the race and smoke variables
birthwt <- birthwt %>% 
  mutate(
  race  = factor(race, levels = c(1, 2, 3), labels=c('White','Black','Other')),
  smoke = factor(smoke,levels = c(0, 1), labels=c('No Smoke', 'Smoke')))

# Graph a histogram of the birth weights bwt using ggplot(birthwt, aes(x=bwt)) + geom_histogram()
# Make separate graphs that denote whether a mother smoked during pregnancy by appending + facet_grid()
# Make your grid of graphs vary with SMOKING STATUS changing vertically, and RACE changing horizontally (that is the formula in facet_grid() should have smoking be the y variable and race as the x).

ggplot(birthwt,aes(x=bwt, fill = smoke)) + 
  geom_histogram(color = "black") +
  labs(x="Birth Weight", y = "Count") +
  facet_grid(rows = vars(race), cols = vars(smoke)) +
  theme_bw() +
  scale_fill_hue(name="Smoking Status",    # Legend label, use darker colors
                     breaks=c("No Smoke", "Smoke"),
                     labels=c("Mothers Who Did Not Smoke", "Mothers Who Smoked"),
                     l=40)  
  
# Remove race from the facet grid
# Add the estimated smooth density using the geom_density() command

ggplot(birthwt,aes(x=bwt, fill = smoke)) +
  geom_histogram(aes(y = after_stat(density)),color = "black", alpha =0.5) + 
  geom_density(alpha = 0.2 , color = "blue") +
  labs(x="Birth Weight", y = "Density") +
  facet_grid(smoke ~ .) + # VERTICAL facet layout vs  (.~ smoke) HORIZONTAL
  theme_bw() +
  scale_fill_hue(name="Smoking Status",    # Legend label, use darker colors
                     breaks=c("No Smoke", "Smoke"),
                     labels=c("Mothers Who Did Not Smoke", "Mothers Who Smoked"),
                     l=40)  
```

```{r echo = FALSE, warning=FALSE}

data(ChickWeight)


# Produce a separate scatter plot of weight vs age for each chick. Use color to distinguish the four different Diet treatments.

ggplot(ChickWeight, aes(x =Time, y = weight , group=Chick , color = Diet)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ Diet) +
  labs(x="Age") +
  theme_classic() +
  scale_color_hue(name="Diet",    # Legend label, use darker colors
                     breaks=c("1", "2", "3", "4"),
                     labels=c("Standard Feed", "High Protein", "Vitamin Supplement", "Enhanced Growth" ),
                     l=40)  

