---
title: "Analysis"
output: prettydoc::html_pretty #flexdashboard::flex_dashboard
date: "2026-01-31"
editor_options: 
  chunk_output_type: console # display chunk output in console style instead of inline notebook style.
---

```{r setup, include=FALSE, message=FALSE}  
knitr::opts_chunk$set(echo = FALSE)  
# Global chunk option:
# settings defined here apply to all subsequent chunks unless overridden.
# e.g echo = FALSE hides the R code and shows only the output/results. for all chunks
# message = FALSE , to avoid package startup messages

if (!require(pacman))
  install.packages("pacman") # if pacman doesnt exist on this machine install it
pacman::p_load(
  tidyverse, #meta-pakage , it loads other packages inside of it
  inspectdf,
  plotly,
  janitor,
  visdat,
  esquisse,
  here,
  readxl, 
  dplyr,
  flextable,
  gt,
  reactable
   )

# load_data

yaounde_data <- read_excel(here("data/yaounde_data.xlsx"))
ebola_sierra_leone <- read.csv(here("data", "ebola_sierra_leone.csv"))
yaounde_data
ebola_sierra_leone
```

# Sierra_leone Ebola Cases by District

**This table shows counts of cases by district:**

```{r District_summary, echo=FALSE, warning=FALSE}
district_tab <- janitor::tabyl(ebola_sierra_leone,district) # creates a table summary of cases(n) and proportions(%)
district_tab_arranged <- district_tab %>%  # create a new object
  rename(Ebola_cases = n)%>% # rename n to Ebola_cases
  arrange(Ebola_cases)
knitr::kable(district_tab_arranged) # knitr::kable() removes the ugly console-style ## output and gives you a proper HTML table.

#Take district_tab, rename the column n to Ebola_cases, then sort by Ebola_cases, and save the final result into a new object called district_tab_arranged.


```

# Categories Plot

**The visualization below shows the frequency distribution of categorical variables in the Ebola dataset.**

```{r categ_vars_plot, echo=FALSE, fig.width=15,fig.align='left',warning=FALSE}
categ_vars_plot <- show_plot(inspectdf::inspect_cat(ebola_sierra_leone)) +
  labs(title = "Categorical Variables in Ebola Dataset")

plotly::ggplotly(categ_vars_plot)
```

# Sierra_leone Car Analysis

**These are Sierra_leone favorite Cars:**

- Toyota Prado
- Mitsubishi Pajero
- BMW X5

*You can see more different options* [here](https://www.autotrader.co.za/cars-for-sale?gclid=Cj0KCQiA-YvMBhDtARIsAHZuUzJ9DJNuf4XXR8qZAd9I_B0cowXqTwVqu07MkmHhciabZEu1R_Z81_4aAkgPEALw_wcB)

```{r cars_summary, echo=FALSE,fig.align='left', warning=FALSE}
summary_of_cars <- summary(cars)
knitr::kable(summary_of_cars)
reactable::reactable(summary_of_cars) # output other table forms we can use reactable, gt or flextable packages
```

# Car Perfomance Plots

**This is the plot for cars**:

```{r cars_plot, echo=FALSE,fig.width=10,fig.align='left',}
plot(cars)
```
